##PA1_template
##Reproducible Research - Week 2  
  

###Load libraries
```{r libraries}
library(ggplot2)
library(timeDate)
library(knitr)
```

###Read data file activity.zip into data frame "activity"
```{r activity}
mydir <- "RepData_PeerAssessment1"
myfile <- paste(mydir, list.files(mydir, pattern = "zip"), sep = "/")
myfile
activity <- read.csv(unzip(myfile), header = TRUE, sep = ",")
```

###What is mean total number of steps taken per day?
    
####Calculate sum of steps per day
```{r steps_per_day, echo=TRUE}
steps_per_day <- aggregate(steps ~ date, activity, sum)
```

####Plot histogram steps per day
```{r histogram steps_per_day, echo=TRUE}
ggplot(steps_per_day, aes(x = steps)) +
        geom_histogram(col = "black", binwidth = 1000) +
        labs(title = "Steps Taken per Day",
             x = "Number of Steps per Day",
             y = "Count") +
             theme_bw()
```

####Calculate mean and median of steps per day
```{r mean_median, echo=TRUE}
steps_mean <- mean(steps_per_day$steps, na.rm = TRUE)
steps_median <- median(steps_per_day$steps, na.rm = TRUE)
steps_mean
steps_median
```
  
Answer 
Mean: `r as.character(steps_mean)`  
Median: `r as.character(steps_median)`  
  
  
###What is the average daily activity pattern?

####Calculate average number of steps per interval
```{r}
steps_per_interval <- aggregate(activity$steps,
                                by = list(interval = activity$interval),
                                FUN = mean,
                                na.rm = TRUE)

colnames(steps_per_interval) <- c("interval", "steps")
```

####Plot steps_per_interval
```{r steps_per_interval}
ggplot(steps_per_interval, aes(x = interval, y = steps)) +
        geom_line() +
        labs(title = "Average Daily Activity Pattern",
             x = "Interval",
             y = "Number of Steps Average Across All Days") +
             theme_bw()
```

####Which interval contains max of number of steps
```{r}
ind <- which.max(steps_per_interval$steps)
max_interval <- as.integer(steps_per_interval[ind, ])
max_interval[1]
max_interval[2]
```
  
Answer  
Interval: `r max_interval[1]`, Steps: `r max_interval[2]`  

  
###Imputing missing values

```{r}
count_na <- sum(is.na(activity$steps))
count_na
```

Selected strategy: replace NAs with mean(steps) per interval accross all days  
We already have the dataset steps_per_interval.

####Merge the original data frame with steps_per_interval (outer join)  
```{r}
activity_mrg <- merge(activity, steps_per_interval, 
                      by="interval", 
                      suffixes = c(".x", ".y") , 
                      all = TRUE)
```

####Now, we replace NAs in steps.x with values of steps.y
```{r}
activity_mrg$steps.x[which(is.na(activity_mrg$steps.x))] <-
        activity_mrg$steps.y[which(is.na(activity_mrg$steps.x))]
```

####Histogram of sum of steps per day
```{r sum of steps per day}
steps_per_day_mrg <- aggregate(steps.x ~ date, activity_mrg, sum)

ggplot(steps_per_day_mrg, aes(x = steps.x)) +
        geom_histogram(color = "black", binwidth = 1000) +
        labs(title = "Number of Steps per Day (NAs filled with mean per interval)",
             x = "Number of Steps per Day",
             y = "Count") +
             theme_bw()
```

####Calculate mean and median with NAs filled
```{r mean and median}
steps_mean_fill <- mean(steps_per_day_mrg$steps.x)
steps_median_fill <- median(steps_per_day_mrg$steps.x)
steps_mean <- steps_mean
steps_median <- steps_median
```

#####After filling NAs:  
Mean fill =  `r as.character(steps_mean_fill)`  
Median fill = `r as.character(steps_median_fill)`  

#####Before filling NAs:  
Mean =  `r as.character(steps_mean)`  
Median = `r as.character(steps_median)`  
  
Answer: After filling missing values, mean value remains unchanged and median matches the value of mean.  
  
  
###Are there differences in activity patterns between weekdays and weekends?
 
####Add a factor column daytype to distinguish between weekdays and weekends  
Use isWekend() function from timeDate library
```{r}
activity_day <- activity

activity_day$daytype <- factor(isWeekend(as.Date(activity$date)),
                               levels = c(FALSE, TRUE),
                               labels = c("weekday", "weekend"))
```

####Aggregate number of steps per interval and daytype
```{r}
steps_per_interval_daytype <- aggregate(steps ~ interval + daytype, 
                                data = activity_day,
                                FUN = mean)
```

####Line plot of steps per interval with facets weekday and weekend
```{r steps per interval daytype}
ggplot(steps_per_interval_daytype, aes(x = interval, y = steps)) +
        geom_line() +
        facet_wrap(~ daytype, nrow = 2, ncol = 1) +
        labs(x = "Interval", y = "Number of Steps") +
        theme_bw()
```

Answer: There is one significant peak of steps during the weekday, probably
related to work activities. During the weekend there is a better distribution
of the number of steps and in general there are more steps per interval.

